---
title: "Revenue Optimization & Cancellation Risk Modelling for Hotel Operations"
author: "Eugene Kiepeeh"
date: "2026-01-24"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(ggridges)
library(scales)
library(patchwork)
```


```{r}

hotels <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-02-11/hotels.csv')

```

# Data Preparation

```{r}
hotels <- hotels |>
  mutate(arrival_date = glue::glue("{arrival_date_year}-{arrival_date_month}-{arrival_date_day_of_month}"),
         arrival_date = parse_date(arrival_date, format = "%Y-%B-%d")) |>
  relocate(arrival_date, .before = is_canceled) |>
  select(-agent, -company)
```

```{r}
# Remove duplicates
hotels_df <- unique(hotels)
```


`lead_time`: how far in advance the booking was made. 
            lead_time = 0 → booked on the same day as arrival
            lead_time = 3 → booked 3 days before arrival
            lead_time = 90 → booked 3 months in advance

# Initial EDA

```{r}
hotels_df |>
  mutate(arrival_date_year = as.factor(arrival_date_year)) |>
  ggplot(aes(x = arrival_date_year, fill = hotel)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of bookings across Hotels",
       x = "Year",
       y = "Number of Bookings",
       fill = "Hotel Type") +
  scale_fill_ordinal() +
  theme_bw() +
  theme(legend.position = "bottom")
```


```{r}
hotels_df |>
  mutate(arrival_date_year = as.factor(arrival_date_year)) |>
  ggplot(aes(x = arrival_date_year, fill = reservation_status)) +
  geom_bar(position = "dodge") +
  labs(title = "Bookings Since 2015 Across All Hotels",
       x = "Year",
       y = "Number of Bookings",
       fill = "Hotel Type") +
  scale_fill_ordinal() +
  facet_wrap(~hotel, scale = "free") +
  theme_bw() +
  theme(legend.position = "bottom")
```


```{r}
library(tidyverse)

business_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#1F2A44"),
    plot.subtitle = element_text(size = 11, color = "#4E5D6C"),
    plot.caption = element_text(size = 9, color = "#6B7280", hjust = 0),
    axis.title = element_text(face = "bold", color = "#1F2A44"),
    axis.text = element_text(color = "#374151"),
    panel.grid.minor = element_blank()
  )

```

```{r}
hotels_df |>
  count(arrival_date) |>
  ggplot(aes(x = arrival_date, y = n)) +
  geom_line(color = "#2CB1BC", linewidth = 1) +
  labs(
    title = "Daily Booking Volume Over Time",
    subtitle = "Tracks overall demand patterns and seasonality in hotel bookings",
    x = "Arrival Date",
    y = "Number of Bookings",
    caption = "Source: Hotel booking records. Daily counts aggregated by arrival date."
  ) +
  business_theme
```

```{r}
hotels_df |>
  filter(adr < 1000) |>
  group_by(arrival_date) |>
  summarise(avgPrice = mean(adr, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = arrival_date, y = avgPrice)) +
  geom_line(color = "#1F2A44", linewidth = 1) +
  labs(
    title = "Average Daily Rate (ADR) Trend",
    subtitle = "Shows pricing movement and potential revenue optimization opportunities",
    x = "Arrival Date",
    y = "Average Daily Rate",
    caption = "Extreme ADR values (>1000) excluded to reduce distortion."
  ) +
  business_theme
```

```{r}
hotels_df |>
  mutate(reservation_status = fct_recode(reservation_status, "Canceled" = "No-Show")) |>
  count(arrival_date, reservation_status) |>
  filter(reservation_status == "Canceled") |>
  ggplot(aes(x = arrival_date, y = n)) +
  geom_line() +
  geom_line(color = "#E5533D", linewidth = 1) +
  labs(
    title = "Daily Booking Cancellations",
    subtitle = "Highlights revenue leakage and operational risk from canceled reservations",
    x = "Arrival Date",
    y = "Number of Cancellations",
    caption = "Includes cancellations and no-shows, grouped by arrival date."
  ) +
  business_theme
```



# Additional Data Cleaning
- the following variables were removed becausse of the lack of immediate usefulness to this project currently.
## Addtional Variables **IMPORTANT

```{r}
clean_hotels_df <- hotels_df |>
  mutate(room_match_flag = ifelse(reserved_room_type == assigned_room_type, "Room Match", "Room Change"),
         total_guest = adults + children + babies,
         family_flag = (children + babies) > 0,
         booking_changes_flag = booking_changes > 0) |>
  select(-days_in_waiting_list, -required_car_parking_spaces, -babies, -children, -booking_changes)

# create new variable (values, revenue related)
clean_hotels_df <- clean_hotels_df |>
  mutate(total_nights = stays_in_weekend_nights + stays_in_week_nights,
         revenue = adr*total_nights, 
         booking_time_flag = case_when( 
                          lead_time <= 7 ~ "0-7", 
                        lead_time >7 & lead_time <=30 ~ "8-30",
         lead_time >30 & lead_time <=90 ~"31-90",
         lead_time >90 ~ "90+"),
         weekend_ratio = round(stays_in_weekend_nights/total_nights,2),
         expected_loss = round(is_canceled*revenue,2),
         booking_value = cut(revenue, 
                             breaks = quantile(revenue, 
                                               probs =seq(0,1,0.25), na.rm = TRUE), labels = c("Low Value", "Moderate Value", "High Value", "Premium Value"), include.lowest = TRUE)) |>
  mutate(weekend_ratio = ifelse(weekend_ratio == "NaN", 0, weekend_ratio),
         family_flag = ifelse(is.na(family_flag),0, family_flag),
         booking_urgency = round(1/(lead_time +1),2),
         sameDay_booking = as.character(ifelse(booking_urgency == 1, 1,0)),
         is_canceled = as.factor(is_canceled))
```

## IO Stuff

```{r}
qty_model <- lm(total_nights ~ adr + arrival_date_month, data = clean_hotels_df)
summary(qty_model)
qty_model2 <- lm(total_nights ~ adr, data = clean_hotels_df)
summary(qty_model2)
```

```{r}
cor(clean_hotels_df$total_nights, clean_hotels_df$adr)
cor(clean_hotels_df$total_nights, clean_hotels_df$adr)
```


# Exploratory Data Analysis

```{r}
p1 <- clean_hotels_df |>
  filter(adr < 400) |>
  ggplot(aes(x = adr)) +
           geom_histogram(fill = "darkgreen", color = "White", bins = 30) +
           geom_vline(xintercept = mean(hotels_df$adr), color = "red") +
  scale_y_continuous(expand = expansion(mult = c(0.0,0.08))) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        title = element_text(face = "bold") 
        ) +
  labs(title = "Distribution of Average Daily Rate Per Booking (July 2015 - Aug 2017)",
       x = "Average Daily Rate($)",
       y = "Number of Bookings")
 
p2 <- clean_hotels_df |>
  filter(revenue <2000) |>
  ggplot(aes(x = revenue)) +
           geom_histogram(fill = "darkgreen", color = "White", bins = 30) +
           geom_vline(xintercept = mean(clean_hotels_df$revenue), color = "red") +
  scale_y_continuous(expand = expansion(mult = c(0.0,0.08))) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        title = element_text(face = "bold")) +
  labs(title = "Distribution of Expected Revenue Per Booking (July 2015- Aug 2017)",
       x = "Revenue Per Booking($)",
       y = "Number of Visits")      

p1 + p2 + plot_layout(nrow = 2)
```

```{r}
final_hotel_bookings |>
  filter(reservation_status == "Check-Out") |>
  group_by(as.factor(adults)) |>
  summarize(n = n(), prop = n/87370) |>
  ggplot(aes(x= reorder(`as.factor(adults)`, prop), y = prop)) +
  geom_col() +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.02))) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

## Revenue

```{r Revenue Distribution}
clean_hotels_df |>
  filter(revenue <2000) |>
  ggplot(aes(x = revenue)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "white") +
  scale_x_continuous(labels = dollar_format(), expand = expansion(mult = c(0.0,0.05))) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(), title = element_text(face = "bold")) +
  labs(title = "Expected Revenue Distribution Per Booking (July 2015 - Aug 2017)", 
       x = "Revenue (USD)",
       y = "Number of Bookings",
       caption = "About 550 Bookings had revenue above 2000 were remove to avoid distortion.")

summary(clean_hotels_df$revenue)
```


```{r Total Expected Revenue + Cancellation}
summary_data <- clean_hotels_df |>
  group_by(is_canceled) |>
  summarise(num = n(),
    totalRevenue = sum(revenue, na.rm = T),
            AvgADR = mean(adr, na.rm = T)) |>
  mutate(is_canceled = fct_relevel(is_canceled, "1","0"),
         is_canceled = fct_recode(is_canceled, "Canceled" = "1", "Completed Stay" = "0")) |>
  mutate(share = totalRevenue / sum(totalRevenue))


summary_data |>
  ggplot(aes(x = is_canceled, y = totalRevenue/1e6, fill = is_canceled)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = paste0("$", round(totalRevenue/1e6,1), "M")),
            vjust = -0.5,
            fontface = "bold",
            size = 5) +
  geom_text(data = summary_data |> filter(is_canceled == "Canceled"),
          aes(label = paste0(percent(share), " of total revenue")),
          vjust = 2,
          color = "white",
          fontface = "bold") +
  scale_fill_manual(values = c(
    "Canceled" = "#D7263D",
    "Completed Stay" = "#1B998B")) +
  scale_y_continuous(labels = dollar_format(suffix = "M"), 
                     expand =  expansion(mult = c(0.0,0.12))) +
  labs(
    title = "Revenue Exposure from Booking Cancellations",
    subtitle = "Comparison of Total Expected Revenue: Completed vs. Canceled Bookings",
    x = NULL,
    y = "Total Revenue (USD, Millions)",
    caption = "Source: Hotel Booking Analytics Project") +
    theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 13),
    axis.text.x = element_text(face = "bold"),
    panel.grid.minor = element_blank())
```

```{r}
rev_at_risk <- summary_data |>
  filter(is_canceled == "Canceled") |>
  pull(totalRevenue)

ggplot(summary_data,
       aes(x = is_canceled, y = totalRevenue/1e6, fill = is_canceled)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = paste0("$", round(totalRevenue/1e6,1), "M")),
            vjust = -0.6,
            fontface = "bold",
            size = 5) +
  geom_text(aes(label = paste0("Bookings: ", comma(num))),
            vjust = 2, color = "white", fontface = "bold", size = 4) +
  geom_text(aes(label = paste0("Avg ADR: ", dollar(round(AvgADR,0)))),
          vjust = 3.5, color = "white", fontface =  "bold", size = 4) +
  annotate("label", x = 1.5,
         y = max(summary_data$totalRevenue/1e6) * 1.15,
         label = paste0("Revenue at Risk: ",
                        dollar(rev_at_risk)), fill = "#FFF3CD", 
         color = "black",fontface = "bold", size = 5) +
  scale_fill_manual(values = c("Canceled" = "#D7263D",
    "Completed Stay" = "#1B998B")) +
  scale_y_continuous(labels = dollar_format(suffix = "M"), 
                     expand =  expansion(mult = c(0.0,0.12))) +
  labs(
    title = "Revenue Exposure from Booking Cancellations",
    subtitle = "Comparison of Total Expected Revenue: Completed vs. Canceled Bookings",
    x = NULL,
    y = "Total Revenue (USD, Millions)",
    caption = "Source: Hotel Booking Analytics Project") + 
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 13),
    axis.text.x = element_text(face = "bold"),
    panel.grid.minor = element_blank())
```


```{r}
waterfall_df <- summary_data |>
  summarise(
    Total_Expected = sum(totalRevenue),
    Lost_Cancellations = totalRevenue[is_canceled == "Canceled"],
    Realized_Revenue = totalRevenue[is_canceled == "Completed Stay"])

waterfall_plot_df <- data.frame(
  Stage = c("Total Expected Revenue",
            "Lost to Cancellations",
            "Net Realized Revenue"),
  Amount = c(waterfall_df$Total_Expected,
             -waterfall_df$Lost_Cancellations,
             waterfall_df$Realized_Revenue))
```


```{r}
ggplot(waterfall_plot_df,
       aes(x = Stage,
           y = Amount,
           fill = Amount > 0)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = dollar(Amount)),
            vjust = ifelse(waterfall_plot_df$Amount > 0, -0.5, 1.5), fontface = "bold") +
  scale_fill_manual(values = c("TRUE" = "#1B998B",
                               "FALSE" = "#D7263D")) +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title = "Revenue Impact of Booking Cancellations",
    subtitle = "From Total Expected Revenue to Net Realized Revenue",
    x = NULL,
    y = "Revenue (USD)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    axis.text.x = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

# Factors Influencing Revenue and Cancellation

```{r Lead Time Distribution}
lead_cancel_df1 <- clean_hotels_df |>
  group_by(booking_time_flag) |>
  summarise(bookings = n(),
            cancel_rate = mean(as.numeric(is_canceled))-1,
            cancelBookings = bookings*cancel_rate,
    .groups = "drop") |>
  mutate(
    booking_time_flag = factor(
      booking_time_flag,
      levels = c("0-7","8-30","31-90","90+")),
    booking_share = bookings / sum(bookings))
```


```{r Lead Time Distribution}
ggplot(lead_cancel_df,
       aes(x = booking_time_flag)) +
  geom_col(aes(y = booking_share),
           fill = "#1B998B",
           width = 0.65, alpha = 1) +
  geom_line(aes(y = cancel_rate, group = 1),
            color = "#D7263D", linewidth = 1.2) +
  geom_point(aes(y = cancel_rate),
             color = "#D7263D", size = 3) +
  geom_text(aes(y = booking_share,
                label = percent(booking_share)),
            vjust = 1.5,
            color = "white",
            fontface = "bold") +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, max(lead_cancel_df$booking_share,
                      lead_cancel_df$cancel_rate-1) * 1.15)) +
  labs(title = "Lead Time Distribution and Cancellation Risk",
    subtitle = "Long-Lead Bookings Represent Largest Demand Share and Higher Cancellation Exposure",
    x = "Lead Time (Days Before Arrival)",
    y = "Proportion of Total Bookings / Cancellation Rate") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", size = 18),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(face = "bold"))
```

```{r}
loss_df <- clean_hotels_df |>
  mutate(is_canceled = as.numeric(as.character(is_canceled))) |>
  group_by(booking_time_flag) |>
  summarise(
    expected_loss = sum(revenue * is_canceled, na.rm = TRUE),
    bookings = n(),
    .groups = "drop") |>
  mutate(
    booking_time_flag = factor(booking_time_flag, levels = c("0-7","8-30","31-90","90+")),
    loss_millions = expected_loss / 1e6,
    loss_share = expected_loss / sum(expected_loss))
```


```{r}
ggplot(loss_df,
       aes(x = booking_time_flag,
           y = loss_millions,
           fill = booking_time_flag)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = paste0("$", round(loss_millions,1), "M")),
            vjust = -0.5,
            fontface = "bold",
            size = 5) +
  scale_y_continuous(labels = dollar_format(suffix = "M"),
    expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = c(
    "0-7" = "#1B998B",
    "8-30" = "#2D3047",
    "31-90" = "#FF9F1C",
    "90+" = "#D7263D")) +
  labs(title = "Expected Revenue Loss by Lead Time Segment",
    subtitle = "Long-Lead Bookings Drive Disproportionate Revenue Exposure",
    x = "Lead Time (Days Before Arrival)",
    y = "Expected Revenue Loss (USD, Millions)",
    caption = "Expected Loss = Revenue × Cancellation Indicator") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 13),
    axis.text = element_text(face = "bold"),
    panel.grid.minor = element_blank())
```
- Instead of only analyzing cancellation rates, I quantified financial exposure by segment to identify where revenue management intervention would have the highest impact.”

# 
```{r}
glm_model <- glm(is_canceled ~ lead_time + adr + stays_in_weekend_nights +
    stays_in_week_nights + market_segment + deposit_type,
    data = clean_hotels_df,
    family = "binomial")


summary(glm_model)
exp(coef(glm_model))
```


# Data Maninpulation: Tabular Analysis
```{r}
# company data
clean_hotel_df |>
  summarise(totalRevenue = sum(revenue, na.rm = T),
            AvgADR = mean(adr, na.rm = T))

clean_hotel_df |>
  group_by(is_canceled) |>
  summarise(num = n(),
    totalRevenue = sum(revenue, na.rm = T),
            AvgADR = mean(adr, na.rm = T)) |>
  mutate(is_canceled = fct_relevel(is_canceled, "1","0")) |>
  ggplot(aes(x = is_canceled, y = totalRevenue/1000000)) +
  geom_col() +
  labs(title = "Total Expected Earnings (all bookings)",
       y = "Total Expected Revenue(in millions)",
       x = "Bookings Status")

clean_hotel_df |>
  group_by(reservation_status) |>
  summarise(num = n(),
    totalRevenue = sum(revenue, na.rm = T),
            AvgADR = mean(adr, na.rm = T)) |>
  mutate(propTotalRevenue = totalRevenue/34454779)

```
-32.1+0.0118 = 32.12% of the revenue is at risk.
- 
```{r}
# Cancellation
clean_hotel_df |>
  filter(is_canceled == 1) |>
  group_by(booking_time_flag) |>
  summarize(numCancels = n(), cancelRate = numCancels/24021, .groups = "drop",
            adrRate = sum(adr, na.rm = T),
            revenue = sum(revenue, na.rm = T))
```
- Long lead-time bookings cancel more!

- Does room assignment reduce cancellations
```{r}
clean_hotels_df |>
  filter(is_canceled == 1) |>
  group_by(room_match_flag) |>
  summarise(numCancels = n(), cancelRate = numCancels/24041)
```
- In fact, the room assignment does not affect a client likelihood of cancelling their booking. 

```{r}
clean_hotel_df |>
  group_by(reserved_room_type) |>
  summarize(numBookings = n(), propBookings = n()/87370,
            revenue = sum(revenue, na.rm = T))
```




-- Special Requests
```{r}
clean_hotels_df |>
  ggplot(aes(x = total_of_special_requests, fill = is_canceled)) +
  geom_bar(position = "dodge")
```

# Customer Segmentation

```{r}
market_segment_df <- clean_hotels_df |>
  group_by(market_segment) |>
  summarize(num = n(),
            revenue = sum(revenue, na.rm = T),
            loss = sum(expected_loss, na.rm = T), 
            Gain = (revenue-loss), .groups = "drop") |>
  mutate(pct_bookings = num/sum(num),
         loss_rate = loss/revenue) |>
  filter(market_segment != "Undefined") |>
  pivot_longer(cols = c(loss:Gain), names_to = "Revenue Type",
              values_to = "Value") |>
  mutate(`Revenue Type` = fct_relevel(`Revenue Type`, "loss", "Gain"))
```


```{r Revenue Gain VS Lost by Segment}
market_segment_df |>
  ggplot(aes(x = reorder(market_segment, num/2),
             y = Value/1e6,
             fill = `Revenue Type`)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  scale_y_continuous(labels = scales::dollar_format(suffix = "M")) +
  scale_fill_manual(values = c(
    "loss" = "#D7263D",
    "Gain" = "#1B998B")) +
  labs(title = "Revenue Generated vs Revenue Retained by Market Segment",
    subtitle = "Online Travel Agency generates highest revenue but also highest loss",
    x = NULL, y = "Revenue (Millions USD)",
    fill = "Revenue Type",
    caption = "Gain = Revenue retained after cancellations") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    plot.title.position = "plot")

```


```{r Revenue Lost by Market Segment}
market_segment_df |>
  ggplot(aes(x = reorder(market_segment, loss_rate/2),
             y = loss_rate/2)) +
  geom_col(fill = "#D7263D", width = 0.7) +
  coord_flip() +
  geom_text(aes(label = scales::percent(loss_rate, accuracy = 0.1)),
            hjust = 0.1, size = 4, color = "white", fontface = "bold") +
  scale_y_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Revenue Loss Rate by Market Segment",
    subtitle = "Online TA and Groups segments show highest revenue exposure",
    x = NULL,
    y = "Revenue Lost (%)",
    caption = "Loss Rate = Canceled Revenue / Total Revenue") +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold"))

```

Why This Is Powerful

From your table:

Online TA:

Revenue = 22.26M

Loss = 9.29M

That’s ~42% revenue at risk 🚨

That is HUGE.

Now you can say:

“Although Online TA drives 59% of bookings, it accounts for the highest revenue leakage.”

That’s a business insight.

```{r}
clean_hotels_df |>
  group_by(market_segment) |>
  summarize(num = n(),
            revenue = sum(revenue, na.rm = T),
            loss = sum(expected_loss, na.rm = T), 
            Gain = (revenue-loss), .groups = "drop") |>
    mutate(pct_bookings = num/sum(num),
         loss_rate = loss/revenue) |>
  filter(market_segment != "Undefined") |>
  ggplot(aes(x = revenue/1e6,
             y = loss_rate,
             size = num,
             color = loss_rate,
             label = market_segment)) +
  geom_point(alpha = 1) +
  geom_text(vjust = -1.2, size = 3.5) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_continuous(labels = scales::dollar_format(suffix = "M")) +
  scale_color_gradient(low = "#2E86C1", high = "#C0392B") +
  scale_size_continuous(range = c(4, 14)) +
  labs(
    title = "Revenue vs Revenue Risk by Market Segment",
    subtitle = "Bubble size represents booking volume",
    x = "Total Revenue (Millions USD)",
    y = "Revenue Loss Rate",
    size = "Total Bookings",
    color = "Loss Rate") +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right")
```
This gives you:

X-axis → revenue importance

Y-axis → riskiness

Bubble size → booking volume

Now segments fall into quadrants:

Segment Type	Strategy
High revenue + high loss rate	Tighten cancellation policy
High revenue + low loss rate	Protect & incentivize
Low revenue + high loss rate	Reconsider segment
Low revenue + low loss rate	Low priority

💡 Insight You Can Already Write

From your numbers:

Online TA:

51,609 bookings

22.2M revenue

9.29M lost

That means:

Nearly half of Online TA revenue is lost due to cancellations.

📌 Business Recommendation

Online TA drives majority of bookings and revenue

However, it has the highest revenue leakage

Suggest:

Stricter deposit policy

Tiered cancellation window

Incentivize direct booking

```{r}
deposit_segment <- clean_hotels_df |>
  group_by(deposit_type) |>
  summarise(nBookings = n(),
            nCancels = mean(as.numeric(is_canceled)),
            revenue = sum(revenue, na.rm = T),
            loss = sum(expected_loss, na.rm = T),
            loss_rate = loss/revenue,
            Gain = revenue - loss,
            .groups = "drop")

sum(clean_hotels_df$revenue)
```


```{r}
deposit_segment_df <- clean_hotels_df |>
  group_by(market_segment, deposit_type) |>
  summarize(
    revenue = sum(revenue, na.rm = TRUE),
    loss = sum(expected_loss, na.rm = TRUE),
    loss_rate = loss / revenue,
    .groups = "drop"
  ) |>
  filter(market_segment != "Undefined")

deposit_segment_df |>
  ggplot(aes(x = deposit_type,
             y = reorder(market_segment, loss_rate),
             fill = loss_rate)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "#EBF5FB", high = "#C0392B",
                      labels = scales::percent_format()) +
  labs(
    title = "Revenue Loss Rate by Market Segment and Deposit Type",
    subtitle = "No deposits significantly reduce revenue leakage",
    x = "Deposit Type",
    y = "Market Segment",
    fill = "Loss Rate"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold"),
    plot.title.position = "plot")

```
This chart shows:
Which segments need stricter deposit policies
If Non Refund reduces cancellation impact



🔥 How These 4 Plots Tell a Story

Plot 1 → Which segments are risky?

Plot 2 → Where is revenue concentration risk?

Plot 3 → Does deposit policy reduce exposure?

Plot 4 → Revenue generated vs retained

Conducted revenue-risk analysis on 87K+ hotel bookings, identifying Online TA segment as contributing 59% of bookings but over 40% revenue exposure; proposed deposit policy optimization to reduce expected loss.

### Customer Type
```{r}
customer_segment_df <- clean_hotels_df |>
  group_by(customer_type) |>
  summarise(tBookings = n(),
            cancel_rate = mean(as.numeric(is_canceled))-1,
            cBookings = tBookings*cancel_rate,
            tRevenue = sum(revenue),
            revPerBooking = tRevenue/tBookings,
            tLoss = sum(expected_loss),
            lossPerBook = tLoss/cBookings,
            Gain = tRevenue - tLoss, .groups = "drop") |>
  mutate(pctTBookings =tBookings/sum(tBookings),
         revenue_share = tRevenue/sum(tRevenue),
         loss_share = tLoss/sum(tLoss),
         gain_share = Gain/(sum(Gain)))
```

```{r}
customer_segment_df |>
  ggplot(aes(x = pctTBookings, y = cancel_rate)) +
  geom_smooth(method = "lm", se = FALSE, color = "#191414") +
  geom_label(aes(label = customer_type), fill = "#1DB954", color = "white") +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0.1,0.08))) +
  scale_x_continuous(labels = scales::percent, expand = expansion(mult = c(0.1,0.1))) +
  labs(title = "Customer Segmentation: Bookings and Cancellation Rate",
       x = "Total Bookings (%)",
       y = "Cancellation Rate (%)") +
  theme_minimal(base_family = "Helvetica") +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "#666666"),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank())
```

```{r}
customer_segment_long <- customer_segment_df |>
  pivot_longer(c(loss_share,gain_share), names_to = "Revenue_share2", values_to = "Value_share") |>
  mutate(Revenue_share2 = fct_relevel(Revenue_share2, "loss_share", "gain_share"),
         Revenue_share2 = fct_recode(Revenue_share2, "Revenue Loss" ="loss_share", "Revenue Gain" = "gain_share"),
         customer_type = fct_reorder(customer_type, tBookings, mean)) 

tf_data <- customer_segment_df |>
  select(customer_type, tLoss, Gain) |>
  pivot_longer(c(tLoss, Gain), names_to = "Revenue1", values_to = "Revenue1_Value")

combined_df <- bind_cols(customer_segment_long, tf_data) |>
  select(-`customer_type...14`) |>
  rename(customer_type = customer_type...1)
```


```{r}
combined_df |>
  ggplot(aes(x = customer_type, y = Value_share, fill = Revenue_share2)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65) +
geom_text(aes(label = paste0("$", round(Revenue1_Value/1e6,1), "M")), position = position_dodge(width = 0.7),
            vjust = -0.5,
            fontface = "bold",
            size = 4) +
  scale_fill_manual(values = c(
    "Revenue Gain" = "#1B998B",
    "Revenue Loss" = "#D7263D")) +
  scale_y_continuous(labels = percent_format(), expand = expansion(mult = c(0,0.12))) +
  labs(
    title = "Revenue Gain and Loss by Customer Type",
    subtitle = "Comparing contribution of gains vs losses across customer segments",
    x = "Customer Segment",
    y = "Relative Shares",
    fill = "Revenue Type",
    caption = "Revenue Gain = bookings without cancellation; Revenue Loss = revenue lost due to cancellations") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 13),
    axis.text.x = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank())
```

```{r}
clean_hotel_df |>
  filter(adr <2000) |>
  ggplot(aes(customer_type, adr)) +
  geom_boxplot(fill = "#1B998B", color = "#191414") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

```

```{r}
clean_hotel_df |>
  count(customer_type, market_segment, sort = TRUE) |>
  filter(market_segment != "Undefined") |>
  ggplot(aes(x = reorder(customer_type,n), y = n, fill = market_segment)) +
  geom_col(position = "fill") +
  labs(title = "How are customers booking with us") +
  scale_fill_brewer(palette = 1, type = "div") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank())
```


```{r}
library(tidytext)
clean_hotel_df |>
  count(customer_type, market_segment, is_canceled, sort = TRUE) |>
  filter(market_segment != "Undefined") |>
  ggplot(aes(x = reorder_within(market_segment,n, customer_type), y = n, fill = is_canceled)) +
  geom_col(position = "fill") +
  coord_flip() +
  scale_x_reordered() +
  facet_wrap(~customer_type, scale = "free") +
  labs(title = "How are customers booking with us") +
  scale_fill_brewer(palette = 7, type = "qual") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank())
```

```{r}
segment_df <- clean_hotel_df |>
  count(customer_type, market_segment, is_canceled) |>
  pivot_wider(names_from = is_canceled, values_from = n, values_fill = 0) |>
  mutate(cancel_rate = `1`/ (`0`+`1`)) |>
  filter(market_segment != "Undefined")


ggplot(segment_df,
       aes(x = reorder(market_segment, cancel_rate),
           y = cancel_rate,
           fill = customer_type)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Cancellation Rate by Market Segment and Customer Type",
    subtitle = "Online travel agents show elevated cancellation behavior",
    x = "Market Segment",
    y = "Cancellation Rate",
    fill = "Customer Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
heat_df <- clean_hotel_df |>
  group_by(customer_type, market_segment) |>
  summarise(
    cancel_rate = mean(as.numeric(is_canceled))-1,
    .groups = "drop"
  ) |>
  filter(market_segment != "Undefined")

ggplot(heat_df,
       aes(x = market_segment,
           y = customer_type,
           fill = cancel_rate)) +
  geom_tile(color = "white") +

  scale_fill_gradient2(
    low = "#1B998B",
    mid = "steelblue",
    high = "#D7263D",
    midpoint = mean(heat_df$cancel_rate),
    labels = scales::percent_format()
  ) +

  labs(
    title = "Cancellation Risk Heatmap",
    subtitle = "High-risk combinations of Customer Type and Market Segment",
    x = "Market Segment",
    y = "Customer Type",
    fill = "Cancel Rate"
  ) +

  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
risk_df <- clean_hotel_df |>
  group_by(customer_type, market_segment) |>
  summarise(
    expected_loss = sum(expected_loss),
    .groups = "drop"
  )

```


```{r}
clean_hotels_df |>
  count(distribution_channel, sort = TRUE)

clean_hotels_df |>
  count(as.factor(is_repeated_guest), sort = TRUE)

clean_hotels_df |>
  count(customer_type, sort = TRUE)

clean_hotel_df |>
  count(market_segment, customer_type, sort = TRUE)

clean_hotels_df |>
  count(deposit_type, sort = TRUE)
```

# Revenue By Date

```{r}
# revenue by year, month
clean_hotel_df |>
  mutate(is_canceled = fct_relevel(is_canceled, "1","0")) |>
  ggplot(aes(x = is_canceled, y = revenue/1000000)) +
  geom_col() +
  facet_wrap(~arrival_date_year)
```


```{r}
clean_hotel_df |>
  mutate(
    is_canceled = fct_relevel(is_canceled, "1","0"),
    arrival_date_month = fct_relevel(
      arrival_date_month,
      "January","February","March","April","May","June",
      "July","August","September","October","November","December"
    )
  ) |>
  group_by(is_canceled, arrival_date_month) |>
  summarise(totalRevenue = sum(revenue)/10000, .groups = "drop") |>
  ggplot(aes(x = arrival_date_month,
             y = totalRevenue,
             color = is_canceled,
             group = is_canceled)) +
  geom_point() +
  geom_line() +
  scale_color_manual(values = c("1" = "red", "0" = "darkgreen"))

```

```{r}
library(dplyr)
library(ggplot2)
library(forcats)
library(scales)

# Prepare data
plot_df <- clean_hotel_df |>
  mutate(
    is_canceled = fct_recode(is_canceled,
                             "Canceled" = "1",
                             "Completed Stay" = "0"),
    arrival_date_month = fct_relevel(
      arrival_date_month,
      "January","February","March","April","May","June",
      "July","August","September","October","November","December")) |>
  group_by(is_canceled, arrival_date_month) |>
  summarise(totalRevenue = sum(revenue), .groups = "drop")

# Identify peak month per category
peak_df <- plot_df |>
  group_by(is_canceled) |>
  slice_max(totalRevenue, n = 1)

# Plot
ggplot(plot_df,
       aes(x = arrival_date_month,
           y = totalRevenue,
           color = is_canceled,
           group = is_canceled)) +

  # Highlight peak months
  #geom_point(data = peak_df, size = 4) +

  # Main lines
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +

  # Label peak values
  geom_text(data = peak_df,
            aes(label = dollar(totalRevenue)),
            vjust = -1,
            fontface = "bold",
            show.legend = FALSE) +

  scale_y_continuous(labels = dollar_format(), expand = expansion(mult = c(0.01,0.21))) +
  scale_color_manual(values = c(
    "Canceled" = "#D7263D",
    "Completed Stay" = "#1B998B"
  )) +

  labs(
    title = "Monthly Revenue Trends",
    subtitle = "Comparison of Completed Stays vs. Cancellations",
    x = "Arrival Month",
    y = "Total Revenue",
    color = "Booking Status",
    caption = "Revenue totals shown in USD"
  ) +

  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    panel.grid.minor = element_blank()
  )

```

- Probability of cancellation by revenue class

```{r skip}
library(dplyr)
library(ggplot2)
library(forcats)
library(scales)

plot_df2 <- clean_hotel_df |>
  mutate(
    is_canceled = fct_recode(is_canceled,
                             "Canceled" = "1",
                             "Completed Stay" = "0"),
    arrival_date_month = factor(
      arrival_date_month,
      levels = month.name
    )
  ) |>
  group_by(arrival_date_year, is_canceled, arrival_date_month) |>
  summarise(totalRevenue = sum(revenue), .groups = "drop")

plot_df2 <- plot_df2 |>
  arrange(is_canceled, arrival_date_month, arrival_date_year) |>
  group_by(is_canceled, arrival_date_month) |>
  mutate(
    YoY_growth = (totalRevenue / lag(totalRevenue) - 1)
  ) |>
  ungroup()

peak_df2 <- plot_df2 |>
  group_by(arrival_date_year, is_canceled) |>
  slice_max(totalRevenue, n = 1)

ggplot(plot_df2,
       aes(x = arrival_date_month,
           y = totalRevenue,
           color = factor(arrival_date_year),
           group = arrival_date_year)) +

  geom_line(size = 1.2) +
  geom_point(size = 2) +

  # Highlight yearly peak
  geom_point(data = peak_df2, size = 4, show.legend = FALSE) +
  
  geom_text(
  data = plot_df2 |> filter(arrival_date_year == 2017 & !is.na(YoY_growth)) |>
    slice_max(YoY_growth, n = 1),
  aes(label = percent(YoY_growth)),
  vjust = -1,
  fontface = "bold",
  show.legend = FALSE
) +


  # Facet by booking status
  facet_wrap(~ is_canceled, ncol = 1, scale = "free_y") +

  scale_y_continuous(labels = dollar_format()) +

  scale_color_manual(values = c(
    "2015" = "#7A7A7A",
    "2016" = "#1B998B",
    "2017" = "#D7263D"
  )) +

  labs(
    title = "Monthly Revenue Trends with Year-over-Year Comparison (2015–2017)",
    subtitle = "Revenue Growth Patterns by Booking Status",
    x = "Arrival Month",
    y = "Total Revenue (USD)",
    color = "Year",
    caption = "Peaks highlighted per year"
  ) +

  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 12)
  )

```

--- Weekend revenue
```{r}
clean_hotel_df |>
  mutate(weekendRevenue = round(revenue*weekend_ratio,2)) |>
  summarize(totalRevenue = sum(revenue), 
            weekendRevenue =sum(weekendRevenue)) |>
  pivot_longer(cols = c(1:2), 
               names_to = "Revenue Type", 
               values_to = "Total")
```
- A lot to go this hotel prolly go for work trip.

```{r}
clean_hotel_df |>
  filter(is_canceled == 0) |>
  ggplot(aes(x = total_nights)) +
  geom_histogram()

clean_hotel_df |>
  group_by(sameDay_booking) |>
  summarise(revenue = sum(revenue, na.rm = T))

clean_hotel_df |>
  group_by(booking_value) |>
  summarise(expectedLoss = sum(expected_loss, na.rm = T))

clean_hotel_df |>
  group_by(booking_value, is_canceled) |>
  summarise(numCancels = n())
```


# Machine Learning Data Prep
```{r}
clean_hotels_df

clean_hotels_df1 <- clean_hotels_df |>
  mutate(day_of_week = wday(arrival_date, label = TRUE, abbr = TRUE))

clean_hotels_df1 |>
  count(day_of_week) |>
  ggplot(aes(x = day_of_week, y = n)) +
  geom_col()

clean_hotels_df1 |>
  count(arrival_date_week_number) |>
  ggplot(aes(x = reorder(arrival_date_week_number, arrival_date_week_number), 
             y = n)) +
  geom_col() +
  coord_flip()
```


```{r}
hotelData <- clean_hotels_df1 |>
  select(-hotel, -arrival_date, -arrival_date_year, -arrival_date_day_of_month, -country, -reservation_status_date, -adults) 

hotelData <- hotelData |>
  mutate(across(c(arrival_date_month, meal,market_segment, distribution_channel, is_repeated_guest, reserved_room_type, assigned_room_type, deposit_type, customer_type, reservation_status, room_match_flag, family_flag, booking_changes_flag, booking_time_flag, sameDay_booking), as.factor))
```


```{r}
ggplot(clean_hotels_df1 |> filter(adr <2000), aes(x = adr, y = hotel)) +
  geom_boxplot()

ggplot(clean_hotels_df1 |> filter(adr <2000), aes(x = adr, y = is_canceled)) +
  geom_boxplot()
```


```{r}
priceHotel_noCancel <- hotelData |>
  select(-expected_loss, -reservation_status, -booking_value,-room_match_flag,-assigned_room_type,-revenue) |>
  filter(is_canceled != "1") |>
  select(-is_canceled)

# you may change choose to include "room changes" see if it explains something
priceHotel_df2 <- hotelData |>
  select(-expected_loss, -reservation_status, -booking_value,-room_match_flag,-assigned_room_type,-revenue)
  
```

```{r}
# no adr, keep revenue!!
cancelHotel_1 <- hotelData |>
  select(-expected_loss, -reservation_status, -booking_value,-room_match_flag,-assigned_room_type, 
         -adr)

# adr 
cancelHotel_2 <- hotelData |>
  select(-expected_loss, -reservation_status, -booking_value,-room_match_flag,-assigned_room_type,-revenue)

```

```{r}
library(tidyverse)
```

```{r}
library(tidymodels)
library(usemodels)
library(tidyverse)
library(dslabs)
tidymodels_prefer()
library(rpart)
library(rpart.plot)
```

```{r}
set.seed(2456346)

# startify by year if possible
cancel_split <- initial_split(cancelHotel_1, prop = 3/4, strata = is_canceled)
cancelTrain <- training(cancel_split)
cancelTest <- testing(cancel_split)
```


```{r}
# Model Recipe
cancel_recipe <- recipe(is_canceled ~., data = cancelTrain) |>
  update_role(revenue, new_role = "other")

cancel_recipe2 <- recipe(is_canceled ~., data = cancelTrain) |>
  update_role(revenue, new_role = "other") |>
  step_zv(all_numeric_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors())

# Model Specification
cancel_decision <- decision_tree() |>
  set_mode("classification") |>
  set_engine("rpart")

cancel_logit <- logistic_reg() |>
  set_mode("classification") |>
  set_engine("glm")

# Model Worflow
cancel_workflow <- workflow() |>
  add_recipe(cancel_recipe) |>
  add_model(cancel_decision)

cancel_workflow2 <- workflow() |>
  add_recipe(cancel_recipe2) |>
  add_model(cancel_logit)

# Model Fit
cancel_fit <- fit(cancel_workflow, cancelTrain)
cancel_fit2 <- fit(cancel_workflow2, cancelTrain)

# Model Prediction

predict_df <- augment(cancel_fit, cancelTest) |>
  mutate(probLoss = .pred_1*revenue) # probable loss

predict_df2 <- augment(cancel_fit2, cancelTest) |>
  mutate(probLoss = .pred_1*revenue)
```


```{r}
# Model Assessment
predict_df |>
  conf_mat(truth = is_canceled, estimate = .pred_class)

predict_df2 |>
  conf_mat(truth = is_canceled, estimate = .pred_class)
```


```{r}
predict_df |>
  group_by(is_canceled) |>
  metrics(truth = is_canceled, estimate = .pred_class)

predict_df2 |>
  group_by(is_canceled) |>
  metrics(truth = is_canceled, estimate = .pred_class)
```


```{r ROC}
roc_tbl <- predict_df |>
  roc_curve(is_canceled, .pred_1)

roc_tbl2 <- predict_df2 |>
  roc_curve(is_canceled, .pred_1)

autoplot(roc_tbl)
autoplot(roc_tbl2)

predict_df2 |>
  roc_curve(is_canceled, .pred_1) |>
  filter(!.threshold %in% c("-Inf","Int")) |>
  ggplot(aes(x = 1-specificity, y = sensitivity)) +
  geom_point() + 
  labs(caption = "TPR = Sensititvity | TNR = Specificity | 1-sen = Specificity")
```

```{r}
cancel_fit |>
  extract_fit_engine() |>
  rpart.plot(roundint=FALSE)
```

```{r}
predict_df |>
  ggplot(aes(x = probLoss, y = .pred_class, 
             color = .pred_class)) +
  geom_point() +
  theme_minimal()

predict_df2 |>
  drop_na(.pred_class) |>
  ggplot(aes(x = probLoss, y = .pred_class, 
             color = .pred_class)) +
  geom_point() +
  theme_minimal()
```

```{r}
quantile(clean_hotels_df1$revenue)
predict_df |>
  mutate(class = case_when(.pred_1 >= 0.65 & revenue >= 500 ~ "High Risk, High Value", .pred_1 < 0.4 & revenue >= 500 ~ "Low Risk, High Value", TRUE ~ "Others")) |>
  group_by(class) |>
  summarize(totalRevenue = sum(revenue), 
            nClients = n(),
            actual = mean(as.numeric(is_canceled))-1)

predict_df |>
  count(is_canceled)
```

- this shows that if you focus on low risk, high value bookings, our model can save the entity about 3.6 in million because only 25% of those bookings actually canceled based on historical data.

```{r}
library(vip)
vip(cancel_fit) # decision Tree
vip(cancel_fit2) # logistic regression
```

```{r}
clean_hotels_df |> count(market_segment, distribution_channel)
```

