---
title: "Analyzing Canceled Bookings"
author: "Eugene Kiepeeh"
date: "2026-08-21"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(ggridges)
library(scales)
library(patchwork)
```

```{r}
final_hotel_bookings <- read_csv("~/Documents/00 Working Projects/Hotel Data Analytics/Hotel Analytics/Data/Processed_Data/final_hotel_bookings.csv")
```

-----------------
--- Summary Analytics ---

```{r}
sum_cancel <- sum(final_hotel_bookings$is_canceled == 1)

final_hotel_bookings |>
  filter(is_canceled == 1) |>
  group_by(booking_time_flag) |>
  summarize(numCancels = n(), cancelRate = numCancels/sum_cancel, 
            .groups = "drop",
            mean_ADR = mean(adr, na.rm = T),
            expected_rev = sum(revenue, na.rm = T))
```
- Does room assignment reduce cancellations
```{r}
final_hotel_bookings |>
  filter(is_canceled == 1) |>
  group_by(room_match_flag) |>
  summarise(numCancels = n(), cancelRate = numCancels/24041)
```
- In fact, the room assignment does not affect a client likelihood of cancelling their booking. 

# Factors Influencing Revenue and Cancellation

```{r Lead Time Distribution 1}
lead_cancel_df <- final_hotel_bookings |>
  group_by(booking_time_flag) |>
  summarise(bookings = n(),
            cancel_rate = mean(as.numeric(is_canceled))-1,
            cancelBookings = bookings*cancel_rate,
    .groups = "drop") |>
  mutate(
    booking_time_flag = factor(
      booking_time_flag,
      levels = c("0-7","8-30","31-90","90+")),
    booking_share = bookings / sum(bookings))
```


```{r Lead Time Distribution 2}
ggplot(lead_cancel_df,
       aes(x = booking_time_flag)) +
  geom_col(aes(y = booking_share),
           fill = "#1B998B",
           width = 0.65, alpha = 1) +
  geom_line(aes(y = cancel_rate, group = 1),
            color = "#D7263D", linewidth = 1.2) +
  geom_point(aes(y = cancel_rate),
             color = "#D7263D", size = 3) +
  geom_text(aes(y = booking_share,
                label = percent(booking_share)),
            vjust = 1.5, color = "white", fontface = "bold") +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, max(lead_cancel_df$booking_share,
                      lead_cancel_df$cancel_rate-1) * 1.15)) +
  labs(title = "Lead Time Distribution and Cancellation Risk",
    subtitle = "Long-Lead Bookings Represent Largest Demand Share and Higher Cancellation Exposure",
    x = "Lead Time (Days Before Arrival)",
    y = "Proportion of Total Bookings / Cancellation Rate") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", size = 18),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(face = "bold"))
```

```{r}
loss_df <- final_hotel_bookings |>
  #mutate(is_canceled = as.numeric(is_canceled)) |>
  group_by(booking_time_flag) |>
  summarise(
    expected_loss = sum(revenue * is_canceled, na.rm = TRUE),
    bookings = n(),
    .groups = "drop"
  ) |>
  mutate(
    booking_time_flag = factor(booking_time_flag, levels = c("0-7","8-30","31-90","90+")),
    loss_millions = expected_loss / 1e6,
    loss_share = expected_loss / sum(expected_loss))
```


```{r}
ggplot(loss_df,
       aes(x = booking_time_flag,
           y = loss_millions,
           fill = booking_time_flag)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = paste0("$", round(loss_millions,1), "M")),
            vjust = -0.5,
            fontface = "bold",
            size = 5) +
  scale_y_continuous(labels = dollar_format(suffix = "M"),
    expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = c("0-7" = "#1B998B", "8-30" = "#2D3047",
    "31-90" = "#FF9F1C", "90+" = "#D7263D")) +
  labs(title = "Expected Revenue Loss by Lead Time Segment",
    subtitle = "Long-Lead Bookings Drive Disproportionate Revenue Exposure",
    x = "Lead Time (Days Before Arrival)",
    y = "Expected Revenue Loss (USD, Millions)",
    caption = "Expected Loss = Revenue × Cancellation Indicator") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 13),
    axis.text = element_text(face = "bold"),
    panel.grid.minor = element_blank())
```
- Instead of only analyzing cancellation rates, I quantified financial exposure by segment to identify where revenue management intervention would have the highest impact.
